---
title: "WNV Model EDA Highlights"
author: "Andrew Bauman, PhD"
date: "January 28, 2017"
output: github_document
---

# Introduction #

Exploratory Data Analysis for informing West Nile Virus (WNV) probability models.
Selected plots are including, all plots can be found at [WNV_EDA.R]().

```{r global_options, include=FALSE}

knitr::opts_chunk$set(warning=FALSE, message=FALSE)

```

```{r echo=FALSE}
# read training data
train <- read.csv('train.csv') # train data merged with weather data (both stations)
train_station_1 <- read.csv('train_station_1.csv') # station 1 weather only
train_station_2 <- read.csv('train_station_2.csv') # station 2 weather only

# convert staton to factor for use in plotting
train$Station <- factor(train$Station)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load libraries

library(ggplot2)
library(tidyr)
library(dplyr)
library(data.table)
library(RColorBrewer)
library(scales)
source('plotting_tools.R')
```
## Selected Density Pots

Overlaid density plots of Tmin and Tmax by station, where the dashed plot is the mean of the station data. Note that both stations cover the same temperature range, this information is not informative of whether or not to retain data from one station over another or to use the mean. This may be informed by plots temparature vs. WNV rates or through the ML feature selection process.

```{r echo=FALSE}
ggplot(train, aes(x=Tmax, fill=Station)) +
  geom_density(alpha= 0.2) +
  geom_density(aes(x = blended_Tmax), alpha= 0, linetype= 'dashed') +
  my_theme() +
  ggtitle("Density Plot of Tmax by Station")

ggplot(train, aes(x=Tmin, fill=Station)) +
  geom_density(alpha= 0.2) +
  geom_density(aes(x = blended_Tmin), alpha= 0, linetype= 'dashed') +
  my_theme() +
  ggtitle("Density Plot of Tmin by Station")
```

## WNV Rate by Location

The plots below demonstrate a clear relationship between location and the WNV rate.

```{r echo=FALSE, fig.height= 15, fig.width= 12}
ggplot(train_station_1, aes(x=reorder(Block, WnvPresent, function(x){mean(x)}),
                            y= WnvPresent)) +
  geom_bar(stat= "summary", fun.y= "mean", color= "blue",
           width= 0.05) +
  coord_flip() +
  ggtitle("WNV Rate by Block") +
  xlab("Block") +
  ylab("WNV Rate")
```

```{r echo=FALSE, fig.height= 20, fig.width= 12}

ggplot(train_station_1, aes(x=reorder(Street, WnvPresent, function(x){mean(x)}),
                            y= WnvPresent)) +
  geom_bar(stat= "summary", fun.y= "mean", color= "blue", width= 0.05) +
  coord_flip() +
  ggtitle("WNV Rate by Street") +
  xlab("Street") +
  ylab("WNV Rate")
```

## WNV Rate by Species 

This plot demonstrates a relationship between species and wNV presence.

```{r echo=FALSE}
ggplot(train_station_1, aes(x=reorder(Species, WnvPresent, function(x){mean(x)}),
                            y= WnvPresent)) +
  geom_bar(stat= "summary", fun.y= "mean", color= "blue",
           width= 0.1, fill= 'blue') +
  coord_flip() +
  ggtitle("WNV Rate by Species") +
  xlab("Species") +
  ylab("WNV Rate")
```


## WNV Rate by Length of Day

Longer days are associated with WNV presence. 

```{r echo=FALSE}
ggplot(train, aes(x=reorder(DayLength_NearH, WnvPresent, function(x){mean(x)}),
                            y= WnvPresent)) +
  geom_bar(stat= "summary", fun.y= "mean", color= "blue",
           width= 0.1, position= "dodge", fill= 'blue') +
  coord_flip() +
  ggtitle("WNV Rate by Day Length") +
  xlab("Length of Day (hours)") +
  ylab("WNV Rate")
```


## WNV Rate by Sunrise Time 

The number of hours between sunrise and the beginning of the day is predictive
of WNV presence.

```{r echo=FALSE}
ggplot(train, aes(x= cut(train$Sunrise_hours, seq(4,6,0.1), right=FALSE),y= WnvPresent)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Sunrise Hours") +
  xlab("Hours since Midnight") +
  ylab("WNV Rate")
```


## WNV Rate by Temperature Related Variables

WNV presense appears to follow rising temperatures.  The plot for Tmax is particularly interesting as there may be a high and low temperature cluster.

```{r echo=FALSE}
ggplot(train, aes(x= cut(train$Tavg, seq(40,100,5), right=FALSE),y= WnvPresent,
                  color= Station)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Average Temperature") +
  xlab("Celsius") +
  ylab("WNV Rate")
  

ggplot(train, aes(x= cut(train$Tmin, seq(40,100,5), right=FALSE),y= WnvPresent,
                  color= Station)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Minimum Temperature") +
  xlab("Celsius") +
  ylab("WNV Rate")


ggplot(train, aes(x= cut(train$Tmax, seq(40,100,5), right=FALSE),y= WnvPresent,
                  color= Station)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Maximum Temperature") +
  xlab("Celsius") +
  ylab("WNV Rate")
```

Depart, heat, and cool do not appear to be particularly informative.

```{r echo=FALSE}

ggplot(train, aes(x= cut(train$Depart, seq(-12,18,5), right=FALSE),y= WnvPresent)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Depart") +
  xlab("Celsius") +
  ylab("WNV Rate")

ggplot(train, aes(x= cut(train$Heat, seq(0,15,3), right=FALSE),y= WnvPresent)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Heat") +
  xlab("Celsius") +
  ylab("WNV Rate")

ggplot(train, aes(x= cut(train$Cool, seq(0,22,4), right=FALSE),y= WnvPresent)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Cool") +
  xlab("Celsius") +
  ylab("WNV Rate")

```


## WNV Rate by Total Precipitation

A model where lower precipitaion is predictive of WNv pressence is consitent
with the literature for mid-west climate zones.

```{r echo=FALSE}
ggplot(train, aes(x= cut(train$PrecipTotal, seq(0,4,1), right=FALSE),y= WnvPresent)) +
  geom_point(stat= "summary", fun.y= 'mean', size= 2) +
  ggtitle("WNV Rate by Total Precipitation") +
  xlab("Precipitation (inches)") +
  ylab("WNV Rate")
```